---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

Loading necessary libraries:
```{r}
suppressMessages({
    library(readr)
    library(dplyr)
    library(ggplot2)
})
```
Read the data file and create a summarised version per day:
```{r}
activity <- read_csv("activity.zip", col_types = cols(date = col_date(format = "%Y-%m-%d"), 
                                                      interval = col_number(), steps = col_number()))
daily_activity <- activity %>% group_by(date) %>% summarise(steps = sum(steps))
```



## What is mean total number of steps taken per day?
```{r}
ggplot(daily_activity, aes(x = steps)) + 
    geom_histogram(fill = "lightblue", colour = "white", bins = 30, na.rm = TRUE) +
    xlab("Total steps per day")
```

The mean total number of steps per day is
```{r}
mean(daily_activity$steps, na.rm = TRUE)
```

The median number of steps per day is
```{r}
median(daily_activity$steps, na.rm = TRUE)
```



## What is the average daily activity pattern?



## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
